Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2023-01-06 analysis/paper_draft3.Rmd
updated draft text

Relying on historical strike data, however, requires an approach to missingness in the historical data or to the possibility that attorney's strike bias depends on certain trial-level characteristics that vary across trials (e.g., defendant race, charge severity). As a result, there may be incompatibility between plausible values of any attorney's bias parameter based on strike data for an attorney in the current trial and based on historical strike data for that attorney. Simply pooling the the strike data from current and past trials, and analyzing such data together, ignores such incompatibility.

@@

The use of the power prior in this context allows lawyers to explicitly control the assumed degree of incompatibility between the current and historical trial data. Specifically, lower values (closer to 0) of the power prior parameter, $\alpha$, should be selected if incompatibility is expected to be high. Higher values (closer to 1) of the power prior parameter should be selected if incompatibility is expected to be low. Accordingly, the lawyer bringing the _Batson_ challenge, after choosing the value of $\alpha$, must be able to defend that choice as appropriately cooresponding to the likely degree of incompatibility between the current and historical strike data in the particular case.

@@

Statistical methods for _Batson_ and similar challenges aim to help measure how much prospective jurors' race, gender, or other cognizable-class membership affected an attorney's use of peremptory strikes against them. These statistical methods do not aim to  _prove_ bias alone, but rather to lend support to an inferrence of bias and a subsequent _Batson_ challenge.

The Bayesian approach here accounts for available historical strike data when estimating the posterior for the bias parameter. By incorporating the power prior, one can easily assess how sensitive that posterior estimate is to the available historical strike data. With a tool like our prototype app, attorneys can use the approach here in to better detect strike bias for _Batson_ and related challenges. The posterior distribution of the striking attorney's bias parameter can help the lawyer decide in real time whether to make a  _Batson_ challenge at all. If argued, the attorney can introduce that posterior distribution as relevant evidence for presenting a prima facie case and of ultimately proving illegal bias (Figure \@ref(fig:batsonworkflowfig)). To be sure, such evidence alone cannot account for the veracity of the striking attorney's proferred reasons. Besides, _Batson_ procedure allows any type of evidence, not just statistical evidence, for ultimately proving or disproving illegal bias in the particular case. Thus, as with any use of historical strike data, the approach here cannot always determine whether a _Batson_ challenge should succeed or fail.

The approach here also requires more from those who would collect the appropriate historical strike data. For example, given the model we used, we required data on the cognizable-class composition of the prospective jurors not just at the time of the first strike, but at each time either party used a strike.

@@

By their choice of $\alpha$, the lawyer explicitly and transparently denotes the degree of weight placed on the historical strike data in the analysis. This transparency matters, because it thwarts lawyer efforts to strategically select $\alpha$ values to prove or disprove illegal strike bias. Rather, at the key steps of the _Batson_ challenge process (Figure \@ref(fig:batsonworkflowfig)), the lawyers on both sides of a _Batson_ challenge know that the lawyers on the other side can always directly and repeatedly assess how sensitive any bias inference is to the choice of $\alpha$. In any case, the choice of $\alpha$ cannot always determine whether a trial judge finds illegal bias, because no approach to inferring bias based on historical strike data has that ability.

>>>>

Relying on historical strike data, however, requires an approach to missingness in the historical data or to the possibility that attorney's strike bias depends on certain trial-level characteristics that vary across trials (e.g., defendant race, charge severity). As a result, there may be incompatibility between plausible values of any attorney's bias parameter based on strike data for an attorney in the current trial and based on historical strike data for that attorney. Simply pooling the the strike data from current and past trials, and analyzing such data together, ignores such incompatibility, which can lead to inflated type I error and reduced power [@Bennett2020].

@@

The use of the power prior in this context allows lawyers to explicitly control the assumed degree of incompatibility between the current and historical trial data. A commonly adopted practice is to fix the power prior parameter $\alpha$ and assess the sensitivity of results to different values [@Carvalho & Ibrahim]. Specifically, lower values (closer to 0) of $\alpha$ should be selected if incompatibility is expected to be high. Higher values (closer to 1) of the power prior parameter should be selected if incompatibility is expected to be low. Accordingly, the lawyer bringing the _Batson_ challenge, after choosing the value of $\alpha$, must be able to defend that choice as appropriately corresponding to the likely degree of incompatibility between the current and historical strike data in the particular case. Since this overall procedure is to estimate bias to support the bringing of a _Batson_ challenge, the senstivity of bias detection to the value of $\alpha$ is used to evaluate sensitivity of the estimation of bias, not to indicate whether a _Batson_ challenge should succeed.

@@

Statistical methods for _Batson_ and similar challenges aim to help measure how much prospective jurors' race, gender, or other cognizable-class membership affected an attorney's use of peremptory strikes against them. These statistical methods do not aim to  _prove_ bias alone, but rather to lend support to an inference of bias and a subsequent _Batson_ challenge. Previously used methods have either ignored historical trial data, or conducted analyses where current and historical data are merged, with no regard for possible incompatibility between trials. The Bayesian approach presented here accounts for available historical strike data when estimating the posterior for the bias parameter. By incorporating the power prior, one can assess how sensitive that posterior estimate is to the available historical strike data. This is appealing because it makes assumptions about the incompatibility between current and historical trial data an explicit part of the process of identifying bias and the decision of whether to bring a _Batson_ challenge or not.

Moreover, with a tool like our prototype app, attorneys can use the approach presented here to better detect strike bias for _Batson_ and related challenges. The posterior distribution of the striking attorney's bias parameter can help the lawyer decide in real time whether to make a  _Batson_ challenge at all. If argued, the attorney can introduce that posterior distribution as relevant evidence for presenting a prima facie case and of ultimately proving illegal bias (Figure \@ref(fig:batsonworkflowfig)).

To be sure, such evidence alone cannot account for the veracity of the striking attorney's proferred reasons. Besides, _Batson_ procedure allows any type of evidence, not just statistical evidence, for ultimately proving or disproving illegal bias in the particular case. Thus, as with any use of historical strike data, the approach here cannot always determine whether a _Batson_ challenge should succeed or fail. The approach here also requires more from those who would collect the appropriate historical strike data. For example, given the model we used, we required data on the cognizable-class composition of the prospective jurors not just at the time of the first strike, but at each time either party used a strike.

@@

By their choice of $\alpha$, the lawyer explicitly and transparently denotes the degree of weight placed on the historical strike data in the analysis. This transparency matters, because it thwarts lawyer efforts to strategically select $\alpha$ values to prove or disprove illegal strike bias. Rather, at the key steps of the _Batson_ challenge process (Figure \@ref(fig:batsonworkflowfig)), the lawyers on both sides of a _Batson_ challenge know that the lawyers on the other side can always directly and repeatedly assess how sensitive any bias estimation is to the choice of $\alpha$. In any case, the choice of $\alpha$ cannot always determine whether a trial judge finds illegal bias, because no approach to estimating bias based on historical strike data has that ability.

In future work we hope to examine data-driven approached to the selection of $\alpha$; where the degree of incompatibility can be estimated and used to inform the choice of the degree of discounting to apply to historical data. Knowing how to quantify incompatibility and how to specify and appropriate prior for  $\alpha$ is an area of active research [e.g., @Pawel2022]
